Vagueness, Noise, and Communication

### 2014-09-18 paper/02_noisyRMD.tex
diffusion section ready

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their noise-perturbed realizations
$C(\Sstrat)$ and $C(\Rstrat)$. The idea is that $\Sstrat$ and
$\Rstrat$ are what, on average, the idealized, confusion-free behavior
would be, while $C(\Sstrat)$ and $C(\Rstrat)$ are behavioral
strategies that describe the agents' actual noise-perturbed
probabilistic behavior. If we conceive of behavioral strategies
$\Sstrat$ and $\Rstrat$ as row-stochastic matrices, the effect of
state confusability is easily captured by matrix multiplication as:
\begin{align}
  \label{eq:confusion-function}
  C(\Sstrat) & = C \Sstrat &    C(\Rstrat) & = \Rstrat C\,.
\end{align}

The discrete-time replicator diffusion dynamic takes the replicator
dynamic as basic, but factors in the confusion of states at each
update step in a sequential update:
\begin{align*}
  \RDD(\Sstrat) & = C(\RD(\Sstrat)) &   \RDD(\Rstrat) & = C(\RD(\Rstrat)) \,.
\end{align*}
This is equivalent to the following, perhaps more transparent
formulation, given by \citet{Correia2013:The-Bivalent-Tr}:
\begin{align*}
  \RDD(\Sstrat)(\messg \probbar \state_i) & = \sum_{j} C_{ij} \cdot
  \frac{\Sstrat(\messg \probbar \state_j) \cdot
    \EU(\messg \myts ; \myts \state_j,\Rstrat)}
  {\Phi(\state_j,\Sstrat, \Rstrat)} \\
    \RDD(\Rstrat)(\state_i \probbar \messg) & = \sum_{j} C_{ji} \cdot
  \frac{\Rstrat(\state_j \probbar \messg) \cdot
    \EU(\state_j \myts ; \myts \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}

@@

The discrete-time \rmd has the same sequential nature as the \rdd:
first we compute the fitness-driven change according to the standard
replicator dynamic; then we compute the perturbation from
mutation. This becomes clear if we define an independent mutation
function that maps mixed strategies onto the outcome of (one round of)
mutation: 
\begin{align*}
  Q(\Smixed)_i & =  \sum_j  \Smixed_j \cdot
  Q_{ji} &   R(\Rmixed)_i & =  \sum_{j}  \Rmixed_j \cdot
  R_{ji} \,.
\end{align*}
The discrete-time \rmd given above can then be rewritten as:
\begin{align*}
  \RDD(\Smixed) &= Q(\RD(\Smixed)) &   \RDD(\Rmixed) &= Q(\RD(\Rmixed))\,. 
\end{align*}

@@

For example, consider a signaling game with two states and two
messages. Let the confusion matrix be:
\begin{align*}
  C=
  \begin{pmatrix}
    .8 & .2 \\
    .2 & .8 
  \end{pmatrix}\,.
\end{align*}
The resulting mutation matrices are:
\begin{align*}
S^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & 1 & 0 & 0 & 0 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & 0 & 0 & 0 & 1 \cr}
&                    
  R^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & .64 & .16 & .16 & .04 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & .04 & .12 & .12 & .64 \cr}\,.
\end{align*}
Here, a pair like $21$, for example, refers to a pure sender strategy
with $\Spure(\state_1=\messg_2$ and $\Spure(\state_2) =
\messg_1$. Similarly for the receiver.

@@

\begin{theorem}
  \label{thm:sender-eq}
  If $G(\Smixed) = \Sstrat$, then $G(Q^C(\Smixed)) = C(\Sstrat)$.
\end{theorem}

\begin{theorem}
  \label{thm:receiver-eq}
  If $G(\Rmixed) = \Rstrat$, then $G(R^C(\Rmixed)) = C(\Rstrat)$.
\end{theorem}

>>>>

The aggregate effect of confusion of states on behavioral strategies
can be captured in a function that maps behavioral strategies
$\Sstrat$ and $\Rstrat$ to their diffusions $D(\Sstrat)$ and
$D(\Rstrat)$. The idea is that $\Sstrat$ and $\Rstrat$ are what, on
average, the idealized, noise-free behavior would be, while
$D(\Sstrat)$ and $D(\Rstrat)$ are behavioral strategies that describe
the agents' actual noise-perturbed probabilistic behavior. Since the
effect of confusion of states is that behavior at one choice point
percolates to behavior at similar choice points, we speak of
\emph{diffusion of behavior under confusion of states} (see
Section~\ref{sec:iterated-diffusion} for an investigation into this
diffusion process). If we conceive of behavioral strategies $\Sstrat$
and $\Rstrat$ as row-stochastic matrices, the diffusion effect of
state confusability is easily captured by matrix multiplication as a
diffusion function $\Diff_C$ that, based on confusion matrix $C$, maps
behavioral strategies on their diffused realizations:
\begin{align}
  \label{eq:confusion-function}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align}

The discrete-time replicator diffusion dynamic takes the replicator
dynamic as basic, but factors in the confusion of states at each
update step in a sequential update:
\begin{align*}
  \RDD(\Sstrat) & = \Diff_C(\RD(\Sstrat)) &   \RDD(\Rstrat) & = \Diff_C(\RD(\Rstrat)) \,.
\end{align*}
This is equivalent to the following, perhaps more transparent
formulation, given by \citet{Correia2013:The-Bivalent-Tr}:
\begin{align*}
  \RDD(\Sstrat)(\messg \probbar \state_i) & = \sum_{j} C_{ij} \cdot
  \frac{\Sstrat(\messg \probbar \state_j) \cdot
    \EU(\messg \myts ; \myts \state_j,\Rstrat)}
  {\Phi(\state_j,\Sstrat, \Rstrat)} \\
    \RDD(\Rstrat)(\state_i \probbar \messg) & = \sum_{j} C_{ji} \cdot
  \frac{\Rstrat(\state_j \probbar \messg) \cdot
    \EU(\state_j \myts ; \myts \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}

@@

The discrete-time \rmd has the same sequential nature as the \rdd:
first we compute the fitness-driven change according to the standard
replicator dynamic; then we compute the perturbation from
mutation. This becomes clear if we define independent mutation
functions $\Mutate_{Q,R}$ that maps mixed strategies onto the outcome
of (one round of) mutation:
\begin{align*}
  \Mutate_Q(\Smixed)_i & =  \sum_j  \Smixed_j \cdot
  Q_{ji} &   \Mutate_R(\Rmixed)_i & =  \sum_{j}  \Rmixed_j \cdot
  R_{ji} \,.
\end{align*}
The discrete-time \rmd given above can then be rewritten as:
\begin{align*}
  \RDD(\Smixed) &= \Mutate_Q(\RD(\Smixed)) &   \RDD(\Rmixed) &= \Mutate_R(\RD(\Rmixed))\,. 
\end{align*}

@@

For example, consider a signaling game with two states and two
messages. Let the confusion matrix be:
\begin{align*}
  C=
  \begin{pmatrix}
    .8 & .2 \\
    .2 & .8 
  \end{pmatrix}\,.
\end{align*}
The resulting mutation matrices are:
\begin{align*}
Q^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & 1 & 0 & 0 & 0 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & 0 & 0 & 0 & 1 \cr}
&                    
  R^C & = \bordermatrix{ ~ & 11 & 12 & 21 & 22 \cr
                      11 & .64 & .16 & .16 & .04 \cr
                      12 & .16 & .64 & .04 & .16 \cr
                      21 & .16 & .04 & .64 & .16 \cr
                      22 & .04 & .12 & .12 & .64 \cr}\,.
\end{align*}
Here, a pair like $21$, for example, refers to a pure sender strategy
with $\Spure(\state_1=\messg_2$ and $\Spure(\state_2) =
\messg_1$. Similarly for the receiver.

@@

\begin{theorem}
  \label{thm:sender-eq}
  If $G(\Smixed) = \Sstrat$, then $G(\Mutate_{Q^C}(\Smixed)) = \Diff_C(\Sstrat)$.
\end{theorem}

\begin{theorem}
  \label{thm:receiver-eq}
  If $G(\Rmixed) = \Rstrat$, then $G(\Mutate_{R^C}(\Rmixed)) = \Diff_C(\Rstrat)$.
\end{theorem}

### 2014-09-18 paper/03_simulations.tex
diffusion section ready

\subsection{(Iterated) Diffusion}

We have seen that the \rdd can be conceptualized as a two-step
procedure. We first calculate where the replicator dynamic would take
us, and then apply diffusion. Since the latter step is the novel
addition, we should look at diffusion in isolation to understand what
the \rdd does.

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Physical distance space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \alpha) = \expo \left ( -
    \frac{\card{\mystate{i} - \mystate{j}}^s}{ \alpha^2} \right ) \,,
\end{align}
where $\alpha$ is an imprecision parameter. When $\alpha=0$ agents
perfectly discriminate between states; when $\alpha \rightarrow
\infty$ agents cannot discriminate states at
all. Figure~\ref{fig:NosofskySim} gives an impression of
Nosofsky-similarity for different parameter values. Other
formalizations of perceptual similarity are possible, including ones
that allow for different discriminability in different areas of the
state space, but we stick with Nosofsky's similarity function for the
time being, because it is a mathematically simple, yet established
notion in mathematical psychology.

@@

\begin{itemize}
  \item which confusion matrices: normalized Nosofsky (positive matrices)
  \item repeated applications
  \item diffusion leads to a smoothing out of sender and receiver
    behavior to neighboring states
  \item how much depends on the level of confusion (see Figure~\ref{fig:diffusion-SenRec})
  \item in the limit, this will lead sender and receiver to do the
    same thing in each choice point (states for the sender, messages
    for the receiver); this follows from the XYZ theorem
\end{itemize}

@@

  \caption{Iterated diffusion for arbitraty sender (top) and receiver
    strategy (bottom), with 50 states, 2 messages. Each row
    corresponds to a different confusion level $\alpha$, each colum to
    a different number of applications of iterated diffusion.}
  \label{fig:confusion-SenRec}
\end{figure}

\newpage

\subsection{Simulations}

>>>>

The previous section showed that the \rdd can be conceptualized as a
two-step procedure. We first calculate where the replicator dynamic
would take us, and then apply diffusion. Since the latter step is the
novel addition, we should look at diffusion in isolation to understand
what the \rdd does. This is what we will do in
Section~\ref{sec:iterated-diffusion}. Section~\ref{sec:simulations}
\dots does something else \dots

\subsection{(Iterated) Diffusion}
\label{sec:iterated-diffusion}

Confusion of states should intuitively be a function of their
perceptual similarity. Here and in the following, we assume that the
state space of the sim-max game consists of $n \ge 2$ states that are
equally spaced across the unit interval, including $0$ and $1$. We
assume here that the distance $\card{\mystate{i} - \mystate{j}}$ is
the objective, physical similarity between two states $\mystate{i}$
and $\mystate{j}$. Distance in physical space feeds into a perceptual
similarity function, as described by
\citet{Nosofsky1986:Attention-Simil}:
\begin{align}
  \label{eq:NosofskySim}
  \similarity(\mystate{i}, \mystate{j} \myts ; \myts \alpha) = \expo \left ( -
    \frac{\card{\mystate{i} - \mystate{j}}^s}{ \alpha^2} \right ) \,,
\end{align}
where $\alpha$ is an imprecision parameter. When $\alpha=0$ agents
perfectly discriminate between states; when $\alpha \rightarrow
\infty$ agents cannot discriminate states at
all. Figure~\ref{fig:NosofskySim} gives an impression of
Nosofsky-similarity for different parameter values. Other
formalizations of perceptual similarity are possible, including ones
that allow for different discriminability in different areas of the
state space, but we stick with Nosofsky's similarity function for the
time being, because it is a mathematically simple, yet established
notion in mathematical psychology.

@@

Difusion of behavior under confusion of states was defined in
Equation~(\ref{eq:confusion-function}), repeated here:
\begin{align*}
  \Diff_C(\Sstrat) & = C \Sstrat &    \Diff_C(\Rstrat) & = \Rstrat C\,.
\end{align*}
If states are confused on average with a probability proportional to
their similarity, the effect on noise-perturbed behavior is that
behavioral dispositions gradually diffuse along a gradient of
similarity of states as well. Consequently, iterated diffusion leads
to a smoothing out and an eventual equalization of behavioral
strategies, for both the sender and the
receiver. Figure~\ref{fig:confusion-SenRec} shows an arbitrary sender
and receiver strategy after one or several rounds of iterated
application of diffusion, for different values of perceptual
imprecision $\alpha$.

The example in Figure~\ref{fig:confusion-SenRec} shows that iterated
diffusion leads to sender and receiver behavior that is the same at
all choice points. After $n$ steps of diffusion an initial sender
strategy $\Sstrat$ will be $C^n \Sstrat$, and an initial receiver
strategy will be $\Rstrat C^n$. If we assume that $\alpha > 0$ and so
$C$ is a positive row-stochastic matrix (each state could in principle
be confused as any other state, albeit with possibly a very low
probability), it follows from the Perron-Frobenius theorem that the
limit $C^\infty = \lim_{n \rightarrow \infty} C^n$ exists (it is the
Perron-projection) and all of its rows are identical. That is why, in
the limit of iterated diffusion, $C^\infty S$ has identical rows
(messages are sent with the same probability in each state), and $R
C^\infty$ likewise has identical rows (every message is interpreted
equally). All of this is in line with the intuition that diffusion of
behavior under confusion of states iteratively equalizes behavior for
similar states; if all states can be confused for one another in
principle, behavior smoothes out entirely in the limit.

@@

  \caption{Iterated diffusion for one arbitrary sender (top) and one
    receiver strategy (bottom), for a game with 50 states and 2
    messages. Each row corresponds to a different confusion level
    $\alpha$, each colum to a different number of applications of
    iterated diffusion.}
  \label{fig:confusion-SenRec}
\end{figure}

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics.

\subsection{Exploring the replicator diffusion dynamics}
\label{sec:simulations}
